---
title: "DSK805 - Exercise set 1"
author: "Kirveslahti and Khan"
date: "2026-02-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This exercise set is based on Applied Multivariate Statistical Analysis,
6th international edition, Chapter 2, reinforced with a recap of R.

Topics:

- Basic R and linear algebra
- Reading in data in R
- Elementary plotting
- More Linear Algebra

You may use \texttt{R} for any of the exercises, unless explicitly stated
otherwise. I especially recommend using \texttt{R} for the exercises marked
with $\dagger$. I have also added some comments and follow-up questions for
some of the exercises.



### Section 1 - Basic R commands

Below are some \texttt{R} commands that help you get started. You probably are
familiar with these already, but if not, try them out. 

```{r, eval=FALSE}
###############################
# Creating vectors
###############################
x=c(1,2,3,4) # vector in R is created with the c() command, c for 'concatenate'
# This here is a 4-vector, with elements 1,2,3, and 4
len(x) # returns the dimension of the vector, len for 'length'
# Note that this is not the 'length' of a vector described in the book
# NB: The R-language purists insists on using <- instead of the = operator
# to assign values to variables. These have small nuance differences, that
# we dont need to care about.

# One useful vector is the vector that contains all ones. We can create it as 
# follows:
n=10
y=rep(1, n) #rep for 'repeat'

# (With the same logic, I could create a vector of all zeros, all twos etc.)
# Another useful vector is the vector from 1 to n, which is created as follows:
z=c(1:n) # actually the c() is redundant
```



```{r, eval=FALSE}
###############################
# Creating matrices
###############################
# A matrix is created by specifying a vector that contains all the entries of
# the matrix, and the specifying the shape of the matrix. The byrow=TRUE argument
# tells to fill the matrix from left to right, top to bottom, i.e. row by row
# (the default behavior is to do column by column)

A=matrix(c(1,2,3,4,5,6), ncol=3, nrow=2, byrow=TRUE)
B=matrix(c(1,2,3,4,5,6), ncol=3, nrow=2)

###############
# Basic algebra
###############

# I can add vectors as though they were numbers
# and I can also multiply vectos with scalars (single numbers)
# The result is as expected if the vectors agree in their size

x=c(2,3)
y=c(4,5)
x+y #(6, 8)
x+3*y

# Same goes for matrices

A+B
A+2*B

# If the sizes are different, unexpected behavior can occur:

x+1

# Multiplying vectors and matrices
# I can also multiply vectors and matrices
#######################
# WARNING
# the * operator gives elementwise multiplication:
###########################

x*y
A*B # This is seldom useful, we wont need this in this course

# The more typical Matrix (and vector) multiplication is given by the
# %*% - operation

A1=matrix(c(1,1,1,1), ncol=2, nrow=2)
A2=matrix(c(2,1,2,1), ncol=2, nrow=2)

A1%*%A2 # Multiply A2 from left with A1
A2%*%A1 # Multiply A2 from right with A1
# (Or equivalently, A1 from left with A2)

# Note that above two lines give different results - why?

# We can also use %*% with vectors

x%$*%y # Technically speaking, x and y are 2 by 1 matrices, you should not
# be able to multiply them like this. R automatically treats the other one
# as a 1 by 2 matrix, so actually we are computing t(x)%*%y here
# (This is more commonly known as the dot product, very useful operation)

# The last useful basic operation we need is the transpose:

t(A) # The transpose of A

# We can recover the number of rows and columns a matrix A has with

dim(A) # rows first, then columns

## Getting help
# If you see an R command in the course that you don't undestand, please do
# ask about it. If you need quicker help, Google knows a lot, and there is
# also the help-command. This can be very useful if you forget what parameters
# you need for the function

# Example

help(seq) # Get help on the seq function

# Based on the above we can create a sequence of number
seq(from=0, to=1, by=0.01) # numbers from 0 to 1 with 0.01 increments

```

#### Exercise 1 $\dagger$

Let $\boldsymbol{A}$ be the matrix defined in the \texttt{R} code chunk above, i.e.

$$
\boldsymbol{A}= \begin{pmatrix}
1 & 2 & 3 \\
4 & 5 & 6
\end{pmatrix}.
$$

How many rows and columns does $\boldsymbol{A}$ have? How about $\boldsymbol{A}^T$, the transpose of $\boldsymbol{A}$?

Also compute the matrix products $\boldsymbol{A}^T\boldsymbol{A}$ and $\boldsymbol{A}\boldsymbol{A}^T$.

How many rows and columns do these matrices have? Are these matrices symmetric? 



### Section 2 - Reading in Data

In Statistics we work with data, and that data is typically stored somewhere.
In this Section we go over the data basics in \texttt{R}.

Most of the time the data that we read in is in .DAT files, which we read in as
follows

*NB* Make sure to download the data to your working space before!

Most common way we load data is with the \texttt{read.table} function:

```{r, eval=TRUE}
data=read.table('T1-9R.dat')
```

This is the data in the Table 1.9 from the Book, and it contains (outdated) national records for Women's running for different distances.

Useful summaries of the data are given with

```{r, eval=FALSE}
dim(data) # number of rows and columns
str(data) # Breakdown of each column
summary(data) # The 6-number summary of the numeric variables
```

Modern statistics puts high emphasis on visuals. For this course, there are
two that are clearly more useful than the rest.  Thse are the scatter plot,
and the histogram. First the scatter plot:

```{r, eval=TRUE}
plot(data$V2,data$V3)
grid()
```

We can make this graph more elaborate:

```{r, eval=TRUE}
plot(data$V2,data$V3, xlab='100m record (seconds)', ylab='200m record (seconds)', main='Women\'s national track records 100m vs 200m', pch='+', cex=1.5)
points(mean(data$V2),mean(data$V3), col='red', pch='+', cex=2)
legend(11.7,23, legend=c('Sample Mean'), col='red', pch='+')
grid()
```

It is important to add enough explanation to your graphics, but be careful
not to make them too busy.

#### Exercise 2
The graph above shows the scatter plot Women's national track record for 100 and 200 meters. Each point represents a country, and the x and y axes are the records for 100 and 200 meters, respectively.

Comment on the appearance of the plot: Are the two records independent? What does this mean statistically (mathematically), what does it mean concretely? Why are there no points in the lower right corner of the plot, what would that mean?



The second most useful graphical summary is the histogram.


```{r, eval=FALSE}
hist(data$V3)
```

We can again add bells and whistles to it:

```{r, eval=TRUE}
hist(data$V3, col='red', main='Women\'s national track record - 200 meters', xlab='Seconds')
abline(v=mean(data$V3), col='blue', lwd=4)
abline(v=median(data$V3), col='black', lwd=4)
legend(24, 12 , legend=c('sample mean', 'sample median'), fill=c('blue','black'))
```


The main utility of this histogram tool is to get a sense of the univariate
distribution of a variable. For example here we see two clusters, one peaking around
23, the other at 25. This is also visible in the previous scatter plot, but easier to see with a histogram.

We say that the distribution appears *bimodal* (two modes).
This creates potential challenges for inference.

If you are familiar with the \texttt{ggplot2} package and prefer doing graphics
with that, that is also ok.


### Section 3 - More Linear Algebra

In this section we go beyond the basics in linear algebra to a more specific direction needed in statistics.

Let $\boldsymbol{X}$ be an $n \times p$ matrix, $n > p$.

The column rank of matrix $\boldsymbol{X}$ is the number of linearly independent columns.

If such $\boldsymbol{X}$ is of full column rank, the *Gram matrix* $\boldsymbol{X}^T\boldsymbol{X}$ is of full rank, diagonalizable $p \times p$, positive definite matrix

For such matrices, the following holds:

$$
\boldsymbol{X}^T\boldsymbol{X} = \boldsymbol{QDQ}^T,
$$
where $\boldsymbol{D}$ is a diagonal matrix consisting of the eigenvalues of $\boldsymbol{X}^T\boldsymbol{X}$, and
$\boldsymbol{Q}$ is a unitary matrix (i.e. $\boldsymbol{Q}\boldsymbol{Q}^T=\boldsymbol{Q}\boldsymbol{Q}^T=\boldsymbol{I}$) that consists of the eigenvectors of $\boldsymbol{X}$.

This result is known as the **spectral decomposition** and this is perhaps the most important result from linear algebra that we need on this course.

Some more advances matrix algebra commands in \texttt{R} are listed below:

```{r, eval=FALSE}
M=matrix(c(2,1,1,2), ncol=2,nrow=2)
E=eigen(M) # The eigen decomposition of a matrix
E$values # Eigenvalues
E$vectors # Eigenvectors
det(M) # The determinant of M
sum(diag(M)) # The trace of M
solve(M) # M-inverse (if exists)
isSymmetric(M) # Returns true if M is symmetric

# Other decompositions: (nice to know category)
svd(M) # The singular Value Decomposition (General but less needed)
# We won't discuss details of the svd on this course
chol(M) # The Cholesky Decomposition of M
# The Cholesky decomposition is useful in statistics and in sampling
# but conceptually not as important as the eigen decomposition
```


### Exercise 3.1 (Adapted from AMSA 2.7) $\dagger$

Let 

$$
\boldsymbol{A}= \begin{pmatrix}
9 & -2 \\ -2 & 6
\end{pmatrix}.
$$
\begin{itemize}
\item Is $\boldsymbol{A}$ symmetric?
\item Is $\boldsymbol{A}$ positive definite?
\item Find the Eigendecomposition of $\boldsymbol{A}=\boldsymbol{QDQ}^T$. State explicitly what are the eigenvectors and their corresponding eigenvalues.
\item Verify that the product $\boldsymbol{QDQ}^T$ corresponds to $\boldsymbol{A}$
\item Find the inverse $\boldsymbol{A}^{-1}$, and verify that $\boldsymbol{AA}^{-1}=\boldsymbol{I}$ and that $\boldsymbol{A}^{-1}=\boldsymbol{QD}^{-1}\boldsymbol{Q}^T$.
\item Compute the Determinant and Trace of $\boldsymbol{A}$. How might you express these in terms of the eigenvalues?
\end{itemize}

### Exercise 3.2 (AMSA 2.38) $\dagger$

Find the minimal and maximal values of $\frac{\boldsymbol{x}^T\boldsymbol{Ax}}{\boldsymbol{x}^T\boldsymbol{x}}$ for non-zero $\boldsymbol{x}$ where

$$
\boldsymbol{A}= \begin{pmatrix}
13 & -4 & 2 \\ -4 & 13 & -2 \\ 2 & -2 & 10
\end{pmatrix}.
$$

Hint: Eigendecomposition


### Exercise 3.3 (AMSA 2.15) $\dagger$

A quadratic form $\boldsymbol{x}^T\boldsymbol{Ax}$ is said to be positive definite, if $\boldsymbol{x}^T\boldsymbol{Ax}>0$ for all non-zero $\boldsymbol{x} \in \mathbb{R}^n$.

Is the quadratic form $3x_1^2 + 3x_2^2 - 2x_1x_2$ positive definite?


### Exercise 3.4 (AMSA 2.18) $\dagger$

Consider the set of points ($x_1,x_2$) whose 'distances' from the origin are given as

$$
c^2 = 4x_1^2 + 3x_2^2 - 2 \sqrt{2}x_1x_2,
$$
for $c^2=1$ and $c^2=4$. The set of points $(x_1,x_2)$ that satisfy the equation above
are origin-centered ellipses. Find the minor and major axes of these ellipses, and their lengths.

### Exercise 3.5 (Adapted from AMSA 2.18) $\dagger$

For the matrix 

$$
\boldsymbol{A}= \begin{pmatrix}
9 & -2 \\ -2 & 6
\end{pmatrix}
$$
From Exercise 3.1, find a matrix $\boldsymbol{B}$ such that $\boldsymbol{A}=\boldsymbol{B}^T\boldsymbol{B}$.



#### Exercise 3.6 (AMSA 2.24) $\dagger$

(We have not talked about covariance matrices, but for now, a covariance matrix is a symmetric, positive definite matrix)

Let $\boldsymbol{\Sigma}$ be a covariance matrix

$$
\boldsymbol{\Sigma}= \begin{pmatrix}
4 & 0 & 0 \\ 0 & 9 & 0 \\ 0 & 0 & 1
\end{pmatrix}.
$$
Find the inverse of $\boldsymbol{\Sigma}$, as well as the eigenvalues and eigenvectors for both $\boldsymbol{\Sigma}$ and $\boldsymbol{\Sigma}^{-1}$.

### Exercise 3.7 (AMSA 2.21) $\dagger$

Let

$$
\boldsymbol{A}= \begin{pmatrix}
1 & 1 \\
2 & -2 \\
2 & 2
\end{pmatrix}.
$$

Compute the matrices $\boldsymbol{A}^T\boldsymbol{A}$ and $\boldsymbol{AA}^T$, and their eigenvalues.


### Exercise 3.9 - The Cholesky Decomposition $\dagger$

This is an in-depth exercise that you can safely skip if you are short on time. It will only be discussed if time permits.

Here we learn how to do another 'matrix square root', which is conceptually more complicated but simpler to compute.

For reasons that will become obvious later, we are very interested in this course
in positive definite quadratic forms, that is expressions of the type

$$
\boldsymbol{x}^T\boldsymbol{A}\boldsymbol{x},
$$

Assume we have a matrix $\boldsymbol{L}$, and we let $\boldsymbol{y}=\boldsymbol{L}\boldsymbol{x}$.

Then $\boldsymbol{y}^T\boldsymbol{y} = (\boldsymbol{Lx})^T\boldsymbol{Lx} = \boldsymbol{x}^T\boldsymbol{L}^T\boldsymbol{L}\boldsymbol{x}$.

If we can find $\boldsymbol{L}$ such that $\boldsymbol{A}=\boldsymbol{L}^T\boldsymbol{L}$, then we can express $\boldsymbol{x}^T\boldsymbol{A}\boldsymbol{x}$ in a very simple form as sum of squares:

$$
\boldsymbol{x}^T\boldsymbol{A}\boldsymbol{x}=\boldsymbol{x}^T\boldsymbol{L}^T\boldsymbol{L}\boldsymbol{x}=(\boldsymbol{Lx})^T\boldsymbol{Lx}=\boldsymbol{y}^T\boldsymbol{y} = \sum_{i=1}^{n} y_i^2
$$
We can find many matrices $\boldsymbol{L}$ that satisfy the above. We need to set additional restrictions if we want the solution to be unique.

In the book they talk about a square root matrix, and they specify that $\boldsymbol{L}$ needs to be symmetric.

Another way we could get a square root matrix is by setting $\boldsymbol{A}=\boldsymbol{QDQ}^T$, and then define $\boldsymbol{A}^{1/2}$ as $\boldsymbol{QD}^{1/2}\boldsymbol{Q}^T$.

A third way is to choose $\boldsymbol{L}$ to be lower triangular. This has the advantage that computations are very fast.

**The task:**

Try calculating the Cholesky decomposition for the matrices $\boldsymbol{A}^T\boldsymbol{A}$ from Exercise 3.7. Compare it with other means of computing matrix square roots, for example from the eigen decomposition.



