---
title: "DSK805 - Exercise set 2"
author: "Kirveslahti and Khan"
date: "2026-03-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This exercise set is based on Applied Multivariate Statistical Analysis,
6th international edition, Chapters 3 and 2.

Topics:

- Expectation and Variance
- Correlation and Covariance
- Sample vs. population
- Matrix expressions 

You may use \texttt{R} for any of the exercises, unless explicitly stated
otherwise. I especially recommend using \texttt{R} for the exercises marked
with $\dagger$. I have also added some comments and follow-up questions for
some of the exercises.

Below again we list some handy \texttt{R} commands. The ones from last set are in the Appendix.


```{r, eval=FALSE}

###############
# Helpful commands
# Expectation, Variance, Sample and Population
###############
data=read.table('T1-9R.dat') # read in data from file 'T1-9R.dat to variable data
# (This accepts many useful optional parameters -- see help("read.table")
# read.csv is another one, and 'scan'
# (See the help if interested)

colMeans(data) # the mean vector (provided data is n x p, rows denote samples)
#rowMeans(data) (for row means -- practically seldom needed)
cor(data) # Compute the sample variance-covariance matrix
# (this is the unbiased version, i.e. square sum divided by (n-1))
# (var(data) does the same)
cov(data) # Compute the sample correlations
# Also possible to do cov(y1,y2)

as.matrix(data) # Turn a dataframe into a matrix
# (sometimes useful)

####
#Data Wrangling
####
data[,1] # Select the first column from dataframe data
data[,c(1,2)] # Select the first two columns
data[,-1] # Select everything but the first column
data[,-c(1,2)] # Select everything except for the first two columns
# The above also work for the rows, e.g.
data[c(1:5),] # Select the first 5 rows


mean(data[,1]) # The mean of individual variable (assuming its numeric)
var(data[,1]) # Variance of individual variable

# Combining vectors
Z=cbind(z1,z2,z3) #cbind: column bind: bind n-vectors z1,z2 and z3
# to a n by 3 array
# Similarly we can do
#Z=rbind(z1,z2,z3) 
# This binds matrices vertically (row bind) - again less frequent compared to
# cbind


# Visualizing correlations
# This requires an extra package, but the result is worth it
# (you may need to install it first with install.packages('corrplot'))
# (installation is only needed once, and from there just load with library)
library(corrplot)
corrplot(cor(data), method="color")

```

There are two 1-number summaries of a variance-covariance matrix. First is the generalized sample variance, which is the determinant of the variance-covariance matrix. Another summary is the total sample variance, which is the trace of the variance-covariance matrix. In \texttt{R}, we compute them as follows:

```{r, eval=FALSE}
S=matrix(c(1,-1,-1,1), ncol=2,nrow=2)
det(S) # Compute the determinant of S (generalized sample variance)
sum(diag(S)) # The trace of S (total sample variance)

# Alternative way via the eigen-decomposition

tmp=eigen(S)
prod(tmp$values) # The determinant
sum(tmp$values) # The trace

```

### Exercise 1 (AMSA 2.30)

You are given a random vector $\boldsymbol{X}^T=(X_1,X_2,X_3,X_4)$ with mean vector $\boldsymbol{\mu_X}^T=(4,3,2,1)$ and variance-covariance matrix

$$
\begin{pmatrix}
3 & 0 & 2 & 2 \\
0 & 1 & 1 & 0 \\
2 & 1 & 9 & -2 \\
2 & 0 & -2 & 4
\end{pmatrix}.
$$
Write $\boldsymbol{X}^{(1)}=[X_1,X_2]$ and $\boldsymbol{X}^{(2)}=[X_3,X_4]$, and let
$$
\boldsymbol{A}=\begin{pmatrix} 1 & 2 \end{pmatrix},
$$
and
$$
\boldsymbol{B}=\begin{pmatrix} 1 & -2 \\ 2 & -1 \end{pmatrix}.
$$

Find the following:

\newcommand{\E}{\mathbb{E}}

\begin{itemize}
\item[(a)] $\E[\boldsymbol{X}^{(1)}]$
\item[(b)] $\E[\boldsymbol{A} \boldsymbol{X}^{(1)}]$
\item[(c)] $\textrm{Cov}(\boldsymbol{X}^{(1)})$
\item[(d)] $\textrm{Cov}(\boldsymbol{A} \boldsymbol{X}^{(1)})$
\item[(e)] $\E[\boldsymbol{X}^{(2)}]$
\item[(f)] $\E[\boldsymbol{B} \boldsymbol{X}^{(2)}]$
\item[(g)] $\textrm{Cov}(\boldsymbol{X}^{(2)})$
\item[(h)] $\textrm{Cov}(\boldsymbol{B}\boldsymbol{X}^{(2)})$
\item[(i)] $\textrm{Cov}(\boldsymbol{X}^{(1)},\boldsymbol{X}^{(2)})$
\item[(i)] $\textrm{Cov}(\boldsymbol{A}\boldsymbol{X}^{(1)},\boldsymbol{B}\boldsymbol{X}^{(2)})$
\end{itemize}


### Exercise 2 (AMSA 2.27)

Derive the expressions for means and variances of the following linear combinations
of $X_1,X_2$ and $X_3$ in terms of the means, variances and covariances of the random variables $X_1,X_2,X_3$.

\begin{itemize}
\item[(a)] $X_1-2X_2$
\item[(b)] $-X_1+3X_2$
\item[(c)] $X_1+X_2+X_3$
\item[(e)] $X_1+2X_2-X_3$
\item[(f)] $3X_1-4X_2$ with the additional condition that $X_1$ and $X_2$ are independent.
\end{itemize}

### Exercise 3 (Adapted from AMSA 2.41)

Let $\boldsymbol{X}=(X_1,X_2,X_3,X_4)$ be a random vector with mean $\boldsymbol{\mu_X}=(3,2,-2,0)$ and variance-covariance matrix

$$
\boldsymbol{\Sigma}= \begin{pmatrix} 3 & 0 & 0 & 0 \\ 0 & 3 & 0 & 0 \\ 0 & 0 & 3 & 0 \\ 0 & 0 & 0 & 3  \end{pmatrix}.
$$
Let 
$$
\boldsymbol{A} = \begin{pmatrix} 1 & -1 & 0 & 0 \\ 1 & 1 & -2 & 0 \\ 1 & 1 & 1 & -3 \end{pmatrix}.
$$

Find $\E[\boldsymbol{A}\boldsymbol{X}]$ and $\textrm{Cov}(\boldsymbol{A}\boldsymbol{X})$.

### Exercise 4 (Adapted from AMSA 3.1)

Given the data matrix 

$$
\boldsymbol{X}=\begin{pmatrix} 9 & 1 \\ 5 & 3 \\ 1 & 2 \end{pmatrix}.
$$
\begin{itemize}
\item Plot the scatter plot of the data matrix, and put a sample mean in it (2d plot with 3 points and the mean)
\item Compute the matrix of deviations from the mean (a matrix that shows for each row of $X$ the difference between that row and the sample mean). Which of the datapoints is furthest from the mean?
\end{itemize}


### Exercise 5

Consider the data in Table 1.11 from the book. The data is stored in the file 'T1-11.dat'.

Compute the sample mean and sample variance-covariance matrix by means of linear algebra

Verify these against the \texttt{R} commands presented above.



### Exercise 6 (AMSA 3.7) $\dagger$

Recall: $\boldsymbol{x}^T\boldsymbol{A}\boldsymbol{x}$ is called a quadratic form, and it defines some sort of distance.

Sketch the ellipses defined by the following quadratic forms

$$
\big(\boldsymbol{x}-\bar{\boldsymbol{x}}\big)^T\boldsymbol{S^{-1}}\big(\boldsymbol{x}-\bar{\boldsymbol{x}}\big)=1
$$
\begin{itemize}
\item[(a)] $$\boldsymbol{S}=\begin{pmatrix} 5 & 4 \\ 4 & 5 \end{pmatrix}$$
\item[(b)] $$\boldsymbol{S}=\begin{pmatrix} 5 & -4 \\ -4 & 5 \end{pmatrix}$$
\item[(c)] $$\boldsymbol{S}=\begin{pmatrix} 3 & 0 \\ 0 & 3 \end{pmatrix}$$
\end{itemize}

Also note that all these matrices have the same generalized variance (why?) Which one of them is least variable in terms of total sample variance?


### Exercise 7 (AMSA 3.16)

Let $\boldsymbol{V}$ be a random vector with mean $\boldsymbol{\mu_V}$ and variance-covariance matrix $\E[(\boldsymbol{V}-\boldsymbol{\mu_v})(\boldsymbol{V}-\boldsymbol{\mu_v})^T]=\boldsymbol{\Sigma_V}$.

Show that

$$
\E[\boldsymbol{V}\boldsymbol{V}^T]=\boldsymbol{\Sigma_V}+\boldsymbol{\mu_v}\boldsymbol{\mu_v}^T.
$$

### Exercise 8 (Adapted from AMSA 3.18)

The energy consumption for the US states for 2001 was recorded in a table, from which the following summary statistics were computed. The sample mean

$$
\bar{\boldsymbol{x}} = (0.766,0.508,0.437,0.161)
$$
and the sample variance-covariance matrix

$$
\boldsymbol{S}= \begin{pmatrix} 0.856 & 0.635 & 0.173 & 0.096 \\ 0.635 & 0.568 & 0.128 & 0.067 \\ 0.173 & 0.127 & 0.171 & 0.039 \\ 0.096 & 0.067 & 0.039 & 0.043 \end{pmatrix}.
$$
The presented numbers are in $10^{15}$ BTUs, and the variables correspond to the following energy sources:

\begin{itemize}
\item $x_1$: petroleum
\item $x_2$: natural gas
\item $x_3$: hydroelectric power
\item $x_4$: nuclear electric power
\end{itemize}

\begin{itemize}
\item From the summary statistics, determine the sample mean and sample variance for each energy source
\item Determine the sample mean and sample variance for the excess of petroleum over natural gas consumption (that is to say, how more petroleum was used than natural gas, on average, and how variable this difference was.) 
\item Comment on the general correlation structure on a qualitative level. Are the variables correlated? Is this aligned with what common sense might tell?
\end{itemize}



\newpage

# Appendix.

Helpful R commands from previous exercise rounds:



```{r, eval=FALSE}

###############
#LINEAR ALGEBRA
###############

x=c(1,2,3,4) # Create vector

A1=matrix(c(1,1,1,1), ncol=2, nrow=2) #create a matrix
A2=matrix(c(2,1,2,1), ncol=2, nrow=2)

A1%*%A2 # Multiply A2 from left with A1

n=10
y=rep(1, n) #vector of all ones, 10 times

M=matrix(c(2,1,1,2), ncol=2,nrow=2)
E=eigen(M) # The eigen decomposition of a matrix
E$values # Eigenvalues
E$vectors # Eigenvectors
det(M) # The determinant of M
sum(diag(M)) # The trace of M
solve(M) # M-inverse (if exists)
isSymmetric(M) # Returns true if M is symmetric

# Other decompositions:
svd(M) # The singular Value Decomposition (General but less needed)
chol(M) # The Cholesky Decomposition of M

##############
#DATA HANDLING
##############
data=read.table('T1-9R.dat')
dim(data) # number of rows and columns
str(data) # Breakdown of each column
summary(data) # The 6-number summary of the numeric variables
data$V2 # Get the vector representing variable called V2

##########
#Plotting
##########
plot(data$V2,data$V3) #scatter plot of V2 and V3
hist(data$V3) # histogram of V3
```